Free surface computations in mantle convection models

### 2015-07-08 free-surface-paper.tex
Add a bunch of text on implementation

Goal: provide for free surface computations in mantle convection simulations that are
\begin{itemize}
\item robust, stable, accurate
\item dimension independent
\item work with adaptive mesh refinement
\item parallelized
\item agnostic to mesh geometry
\end{itemize}

@@

\section{Implementation in \texttt{Aspect} }
\subsection{Mass conservation}
\begin{itemize}
\item mass conservation requires $\int_\Omega {\bf u \cdot n} = 0$
\item not all mesh advection schemes respect this
\item for short simulations, it is not likely to make a big difference
\item for long simulations there can be significant changes in domain volume
\item finite element projections do the best job of conservation of mass
\end{itemize}
\subsection{Remeshing}
\begin{itemize}
\item ALE does not specify what happens to interior nodes
\item primary goal is preserving mesh regularity
\item we solve a laplace equation for mesh displacements
\item different boundary conditions for the laplace problem for different velocity boundary conditions
\end{itemize}
\subsection{Parallelization}
\begin{itemize}
\item mesh vertex positions are stored in a distributed vector
\item vector is redistributed with p4est on mesh refinement
\end{itemize}

>>>>

@@

\section{Implementation in \texttt{Aspect} }
We have implemented the ability to run free surface simulations in the mantle convection software \texttt{Aspect} \citep{kronbichler2012high}. 
\texttt{Aspect} is designed to be highly flexible and modular, with the ability for user-defined rheologies, geometries, and gravity models. 
The implementation of the free surface, therefore, needs to be general enough to work for many combinations of these models, including those 
which may not have been written yet. In particular, it cannot rely on assumptions regarding the shape of the domain.

Furthermore, \texttt{Aspect} allows for computations in both 2D and 3D, so the implementation must be sufficiently dimension
independent to work in both cases. We implement the free surface within an arbitrary Lagrangian-Eulerian (ALE) formulation
\citep[e.g][]{fullsack1995arbitrary,donea2004encyclopedia}.

\texttt{Aspect} is a parallelized, distributed memory code with adaptive mesh refinement.
The free surface implementation works with these features.  We store the mesh vertex positions 
in a fully distributed vector. This vector is continually updated and redistributed across 
processes upon mesh adaptation (which is handled by the adaptive octree library \texttt{p4est}).
We also provide adaptive refinement indicators based on being near to the free surface or 
when the free surface slope is steep to allow for accurate interface tracking.

\subsection{Remeshing}
In ALE calculations the internal mesh velicity is undetermined.
In general, one wants to smoothly deform the mesh so as to preserve its regularity, 
avoiding inverted or otherwise poorly conditioned cells.
The mesh deformation can be calcluated in many different ways, including algebraic \citep[e.g.][]{thieulot2011fantom} 
and PDE based approaches.

We choose to implement remeshing based on solving Laplace's equation for the mesh velocity.
We solve the equation
\begin{equation}
\nabla^2 \mathbf{u}_{\mathrm{mesh}} = 0
\label{eq:laplacian_smoothing}
\end{equation}
subject to the boundary conditions
\begin{equation}
\begin{aligned}
&\mathbf{u}_\mathrm{mesh} = \left( \textbf{u} \cdot \textbf{n} \right) \textbf{n} & \qquad & \textrm{on } \partial \Omega_{\textrm{free surface}}, \\
&\mathbf{u}_\mathrm{mesh} \cdot \textbf{n} = 0 & \qquad & \textrm{on } \partial \Omega_{\textrm{free slip}}, \\
&\mathbf{u}_\mathrm{mesh} = 0 & \qquad & \textrm{on } \partial \Omega_{\textrm{Dirichlet}}.
\end{aligned}
\label{eq:laplacian_bcs}
\end{equation}
This scheme has the advantage of working for many different domain geometries and combinations of boundary conditions.
For moderate mesh deformation, the mesh stays smooth and well conditioned, though it breaks down for large deformations 
or non-convex domains.

\subsection{Surface advection}
With a deformable domain there is the danger that small errors in free surface motion can
result in poor overall mass conservation in time. At worst, the total volume of the mesh can 
fluctuate significantly over hundreds or thousands of time steps.
Consistency with the Stokes solution requires 
\begin{equation}
\mathbf{u}_{\mathrm{mesh}} \cdot \mathbf{n} = \mathbf{u \cdot n}
\end{equation}

Unfortunately the normal vectors are not well defined on the mesh vertices, which is 
where the mesh velocity is defined. One could instead advect the mesh in the direction 
of the local vertical, or in some weighted average of the cell normals adjacent to a given vertex,
but we have found that these schemes do not necessarily have excellent mass conservation 
properties.

A better approach is to perform a finite element projection of the Stokes velocity 
solution onto the mesh velocity vector. Multiplying the boundary conditions 
\eqref{eq:laplacian_bcs} by a test function $\mathbf{w}$ and integrating over the free
surface part of the boundary, we find

\begin{equation}
\int_{\partial \Omega_{\textrm{free surface}} } \mathbf{w} \cdot \mathbf{u}_\mathrm{mesh} = 
\int_{\partial \Omega_{\textrm{free surface}} } \left( \mathbf{w \cdot n } \right) \left( \mathbf{u \cdot n} \right)
\end{equation}

When discretized, this forms a linear system which can be solved for the mesh velocity at the 
free surface. Of course, a new system to solve is not ideal, but this system, being nonzero 
over only the free surface, is relatively small and easy to solve.

This weak solution to the boundary conditions~\eqref{eq:laplacian_bcs} is able to borrow
the accuracy of the Stokes solve for Equation~\eqref{eq:weak_incompressible}, and we have 
found it to conserve mass more accurately than algebraic techniques for evaluating the mesh-normal velocity.
Similar results were found by \citet{fullsack1995arbitrary}.
